Graphical Tests for Power Comparison of Competing Designs

### 2012-03-31 intro.tex
restructured a bit, filled in missing pieces
tweaking a little the intro

Here we suggest a new approach using recent research on visual
inference \citep{buja:2009,wickham:2010}. In visual inference, real
data plots are considered to be test statistics, and these are
compared with plots of data generated from a null hypothesis using a
lineup. A lineup is similar to the police lineup, and this is where
the name comes from. The suspect (the plot of the real data) is placed
randomly in a field of plots of null data, and an impartial viewer is
asked to identify the plot that is the most different. If the viewer
identifies the suspect it lends statistical significance to the
conclusion that the real data is not consistent with the null
data. Figure \ref{lineup} has an example of a lineup. Overlaid density
plots are used to display the distribution of two groups (blue, red)
of data. We are interested to know whether the real data has a
significant difference between the centers of the two groups. Plot
numbered 13 contains the real data. Did you pick that one? In this
plot the blue group has a flat density whereas the density of red
group is more peaked and concentrated to the left of the blue.

\begin{figure}[htbp]
   \centering
   \includegraphics[height=0.95\linewidth]{density_1_45_15_20_13.png} 
   \caption{Example lineup using density plots: one plot of real data
     embedded with nineteen plots of null data. Which plot is
     most different from the others? (See the text for the answer.)}
   \label{lineup}
\end{figure}

The previous work \citet{buja:2009,wickham:2010} describes the use of
lineups for evaluating discoveries made in exploratory analysis. Here
we describe the use of lineups for evaluating plot design for
communicating findings visually. Different lineups are created, using
the same data, same null plots and positions within the lineup, but
different plot designs. Comparing how long and how accurately the
viewers take to report the feature of interest will assess which
design is better for the task. To investigate the benefit of lineups
for plot design evaluation we have conducted an experiment that also
contains simulated data, where the data is constructed to match and
expand on the features of interest in the real data. For example, in a
simple situation, where the discovered feature of interest is a
difference between the centers of two distributions, simulated data
would be constructed roughly matching the distribution of the real
data that varies the distance between centers, and also spread and
sample size. Lineups of these simulated data (embedded with simulated
null data plots) are generated and evaluated along with the lineups of
real data. The simulation study provides a backdrop for the real data,
which enables the broad applicability to be studied and a gauge for
the signal strength of the pattern in the real data.

@@

Lineups provide a way to statistically quantify the significance of the finding \citet{majumder:2011}: 
think of the lineup as a set of $m$ plots, one of which is the real data. 
The probability that an observer picks the real data plot just by chance, i.e. when it is really *not* different from the other null plots, is $1/m$. 
If an observer is able to identify the data plot from the lineup, we reject the null hypothesis that the null plot is not different from the data.  This sets  our Type I error rate  $\alpha$ at a level of $1/m$. (In Figure \ref{lineup}, $m=20$)

When there are multiple independent observers ($n$) we have more freedom in setting  the significance level: assume that $x$ out of those $n$ observers picked the data plot. Let $X$ be the corresponding random variable, i.e. $X$ = \# times out of $n$ independent repetitions that  the data is picked from the lineup. Under the null hypothesis, $X$ has a Binomial distribution: $X \sim B_{n, 1/m}$.
We can then compute the $p$ value of a lineup as the probability to have $x$ or more observers picking the data plot (under the assumption that the null hypothesis is true, i.e. the plot is not different):
\[
p\text{-value} = P(X \ge x \mid H_0) =   1 - B_{n, 1/m} (x-1).
\]

When comparing different tests of the same quantity, we consider that test better, that has greater {\it power}.
The power of a test is the probability to reject the null hypothesis -- in a lineup this is the probability that an observer identifies the plot of the real data. 
Analytically, power is usually difficult to calculate because it is requires specific use of an alternative hypothesis to calculate the probability.  Work in \citet{niladri:2012} addresses this to some extent with measures on the quality of a lineup, how different numerically, as best it can be calculated, the real data plot is from the null plots.

@@

Let $n$ be the number of independent observers and $x_i$ the number of observers who picked plot $i$, $i \in \{1, ..., m\}$, from the lineup. Then $(x_1, x_2, ..., x_m)$ follows a multinomial distribution Mult$_{\pi_1, \pi_2, ..., \pi_m}(x_1, x_2, ..., x_m)$ with $\sum_i \pi_i = 1$, where $\pi_i$ is the probability that plot $i$ is picked by an observer, which we can estimate as $\widehat{\pi_i} = x_i/n$.

The power of a lineup can therefore be estimated as the ratio of correct identifications $x$ out of $n$ viewings.

@@

The power of lineups is used in our new work to evaluate the effectiveness of different plot designs. This paper describes two examples where we had real data analysis problems and decisions to make in order to communicate the results. Section \ref{designs} describes the process for comparing designs. Section \ref{setup} and describe the two data analysis problems and the experiments conducted to evaluate the plots designs. Section \ref{results} presents the findings of the experiments, and Section \ref{conclusions} presents conclusions and suggestions for future use.

@@

XXXX should we include nullabor code?

>>>>

Here we suggest a new approach using recent research on visual inference \citep{buja:2009,wickham:2010}. In visual inference, data plots are considered to be test statistics, and these are compared with plots of data generated from a null hypothesis using a lineup. The null hypothesis underlying a lineup is that there is no real structure visible in the data plot, that any patterns seen are consistent with randomness, or from a known model. If the null hypothesis is true then the plot of the data will not be distinguishable form the plots of null data. A lineup is similar to the police lineup, and this is where the name comes from. The suspect (the plot of the data) is placed randomly in a field of plots of null data, and an impartial viewer is asked to identify the plot that is the most different. If the viewer identifies the suspect it lends statistical significance to the conclusion that the data is not consistent with the null data. Figure \ref{lineup} has an example of a lineup. Overlaid density plots are used to display the distribution of two groups (blue, red) of data. We are interested to know whether the data has a significant difference between the centers of the two groups. Plot numbered 13 contains the data. Did you pick that one? In this plot the blue group has a flat density whereas the density of red group is more peaked and concentrated to the left of the blue.

\begin{figure}[htbp]
   \centering
   \includegraphics[height=0.95\linewidth]{density_1_45_15_20_13.png} 
   \caption{Example lineup using density plots: one plot of data
     embedded with nineteen plots of null data. Which plot is
     most different from the others? (See the text for the answer.)}
   \label{lineup}
\end{figure}

The previous work \citet{buja:2009,wickham:2010} describes the use of lineups for evaluating discoveries made in exploratory analysis. Here we describe the use of lineups for evaluating plot design for communicating findings visually. Different lineups are created, using the same data, same null plots and positions within the lineup, but different plot designs. Comparing how long and how accurately the viewers take to report the feature of interest will assess which design is better for the task. To investigate the benefit of lineups for plot design evaluation we have conducted an experiment that also contains simulated data, where the data is constructed to match and expand on the features of interest in the data. For example, in a simple situation, where the discovered feature of interest is a difference between the centers of two distributions, simulated data would be constructed roughly matching the distribution of the data that varies the distance between centers, and also spread and sample size. Lineups of these simulated data (embedded with simulated null data plots) are generated and evaluated along with the lineups of data. The simulation study provides a backdrop for the real data, which enables the broad applicability to be studied and a gauge for the strength of the pattern in the data.

@@

Lineups provide a way to statistically quantify the significance of the finding \citet{majumder:2011}: think of the lineup as a set of $m$ plots, one of which is the data. (In Figure \ref{lineup}, $m=20$.)  The probability that an observer picks the data plot just by chance, i.e. when it is really *not* different from the other null plots, is $1/m$.  If an observer is able to identify the data plot from the lineup, we reject the null hypothesis.  This sets our Type I error rate, $\alpha$, at a level of $1/m$.

When there are multiple independent observers ($n$) we have more freedom in setting  the significance level: assume that $x$ out of those $n$ observers picked the data plot. Let $X$ be the corresponding random variable, i.e. $X$ = \# times out of $n$ independent repetitions that  the data plot is picked from the lineup. Under the null hypothesis, $X$ has a Binomial distribution: $X \sim B_{n, 1/m}$. We can then compute the $p$-value of a lineup as the probability to have $x$ or more observers picking the data plot (under the assumption that the null hypothesis is true, i.e. the plot is not different):
\[
p\text{-value} = P(X \ge x \mid H_0) =   1 - B_{n, 1/m} (x-1).
\]

When comparing different tests of the same quantity, we consider that
test better if it has greater {\it power}.  The power of a test is the
probability to reject the null hypothesis, irrespective of whether it
is true or false -- in a lineup this is the probability that an
observer identifies the plot of the real data.  Analytically, power is
usually difficult to calculate because it is requires specific use of
an alternative hypothesis to calculate the probability.  Work in
\citet{niladri:2012} addresses this to some extent with measures on
the quality of a lineup, how numerically different, as best it can be
calculated, the data plot is from the null plots.

@@

\begin{quote}
Let $n$ be the number of independent observers and $x_i$ the number of observers who picked plot $i$, $i \in \{1, ..., m\}$, from the lineup. Then $(x_1, x_2, ..., x_m)$ follows a multinomial distribution Mult$_{\pi_1, \pi_2, ..., \pi_m}(x_1, x_2, ..., x_m)$ with $\sum_i \pi_i = 1$, where $\pi_i$ is the probability that plot $i$ is picked by an observer, which we can estimate as $\widehat{\pi_i} = x_i/n$. 
\end{quote}

\noindent The power of a lineup can therefore be estimated as the ratio of correct identifications $x$ out of $n$ viewings. (More details on these derivations can be found in \citet{majumder:2012}.)

@@

The power of lineups is used in the work presented here to evaluate the effectiveness of different plot designs. This paper describes two examples where we had real data analysis problems and decisions to make in order to communicate the results. Section \ref{designs} describes the process for comparing designs. Section \ref{setup} and describe the two data analysis problems and the experiments conducted to evaluate the plots designs. Section \ref{results} presents the findings of the experiments, and Section \ref{conclusions} presents conclusions and suggestions for future use.

@@
